MS_Discussion

### 2024-03-19 Markdown/Manuscript/Child_RMDs/Clippings.Rmd
v4

<!-- We have designed this framework to enable ethical computational implementations of health economic models that are both open-source (for transparency and reusability) and modular (for reusability and updatability). The framework aims to support the standardization and interoperability that are core enablers of implementing a modular and open source approaches.  -->

@@

Modular models provide an opportunity for multiple modelling teams to contribute to, test and reuse models. To enhance this capacity, modular models may be implemented as open source projects that give others liberal permissions to access and use model source code and data [@pan2021modular; @barros2023empowering; @frazier2022robust]. Although there appears to be in principle support from many health economists for greater use of open source health economic models (OSHEMs)  [@Pouwels2022], actual implementations are rare [@Emerson2019; @Michalczyk2018; @Feenstra2022]. Barriers to adoption of OSHEMs include concerns about intellectual property, confidentiality, model misuse and the resources required to support open source implementations [@Pouwels2022; @Wu2019]. Adherence to good practice guidance is an essential requirement for healthcare modelling [@Erdemir2020], but guidelines for implementing OSHEMs remain scarce, piecemeal and need improving [@Sampson2019].

Our interest in modular and open source approaches developed when we began seeking am appropriate framework for undertaking and validly synthesising diverse types of economic research in mental health. Mental disorders impose high health, social and economic burdens worldwide [@RN8;@GBD2019]. Much of this burden is potentially avertable [@RN25], but poorly financed and organised mental health systems are ill-equipped for this challenge [@RN22;@RN23]. A substantial economic literature already exists to assess the affordability and value for money of mental health interventions [@RN34]. This economic evaluation work is an essential prerequisite for improving allocative efficiency in mental health, but could be of greater value to systems planners if integrated with a broader program of economic research.

@@

Standardisation is an essential enabler of automation, collaboration, interoperability and transferability in modular and open source approaches. An early requirement to implement ready4 is therefore to identify an appropriate set of standards. However, we are not aware of any consolidated source of guidance about recommended standards for MOSHEMs. Once identified, the standards specified for ready4 will need to be implemented by software, some of which will require custom development.

@@

### Standards for a reusable MOSHEM
```{r echo = F}
gpl_pc_dbl <- get_license_share("GPL")
```
To make model code and data widely re-usable by others, it is important to provide users with appropriate and explicit permissions. In the context of open source models, there are two broad categories of licensing options. Some guidance strongly recommends the use of permissive licensing [@Wilson_2017] that provides users with great flexibility as to the purposes (including commercial) for which the content could be re-used. An alternative approach is to use copyleft licenses [@copyleft2022] that can require content users to distribute any derivative works they create under similar open source arrangements. For code, it may be appropriate to adopt the prevailing open source licensing practice within the programming language being used. For data, it may not be sufficient to simply choose between a permissive license like the Public Domain Dedication (CC0) [@cc02022] or a copyleft option such as the Attribution-Share Alike (CC-BY-SA) [@bysa2022]. In addition to ensuring that data is ethically appropriate for disseminate in open access repositories, responsible custodianship of some de-identified or aggregated data may involve using or adapting template terms of use [@sampleterms2022] which have a number of ethical clauses (for example, prohibiting efforts to re-identify research participants).

@@

Statistical models are a common output of health economic evaluations, but they are often not reported in a format that enables others to confidently and reliably re-use them for out of sample prediction [@Kearns2013]. Open source approaches can help address this by disseminating code artefacts that enable easy and appropriate use of a statistical model to make predictions with new data. However, great care must be exercised when publicly releasing model artefacts derived from data on human subjects as they may by default embed a copy of the source dataset. Sensitive dataset copies must therefore be replaced (for example, with synthetic data) and the amended artefact's predictive performance then retested before any public release. Another way to make MOSHEMs easier to use is to develop simple user-interfaces for non-technical users.

@@

However, having features that facilitate accountability, reuse and updating is not the same as being accountable, reused and updated. If diverse groups of stakeholders do not review model components, suggest improvements and develop alternatives, then little progress is made towards enhancing model legitimacy. Similarly, making code and data publicly available does not guarantee that others will know of the existence of these tools, trust their validity and find them easy to use. Without reuse, errors in model artefacts are more likely to remain undetected. Even when errors are detected, they still need to be fixed, but maintaining code and data requires ongoing resourcing through a combination of centralised infrastructure and an active open source community.

@@

## Implications for economic modelling in mental health
Open source approaches have been recommended to help develop the mental health modelling field [@RN73] but only one mental health related model (in Alcohol Use Disorder [@Basu2018]) is currently indexed in the Open Source Models Clearinghouse [@OSMC_20xx;@Emerson2019]. We are aware of just one other open source mental health model - a reference model in Major Depressive Disorder - that is currently in development [@IVIMDD2022]. Of the known barriers to adoption of open source models by health economists (including issues like intellectual property and confidentiality [@Pouwels2022; @Wu2019]), our experience suggests that the biggest challenges may be the enormous effort required to first prepare model code and data for public release in formats that facilitate appropriate reuse by third parties and to then maintain and continually improve potentially large numbers of digital artefacts.

Automated tools such as those we developed in our modelling toolkit can help reduce the burden associated with some of these tasks. However, we think the current low rates of adoption by health economists of open source approaches [@Emerson2019; @Michalczyk2018; @Feenstra2022] (which in turn facilitate the collaboration that make modular models more attractive) will only change slowly unless there is significant and strategic investments made by research funders. Currently, incentive structures for health economists do not promote the dedication of large quantities of time to enable peers to reuse their work.<!-- MH Research priorities (WP3 - Research, WP6 - Economic) https://doi.org/10.1016/S2215-0366(15)00332-6 -->

Reducing waste in research is a responsibility of research funders [@chalmers2014increase] and the poor reproducibility [@Jalali2021; @McManus2019; @Bermejo2017], limited reusability [@Emerson2019; @Michalczyk2018; @Feenstra2022] and uncertain validity [@Ghabri2019; @Radeva2020] of health economic models is wasteful. Approximately 4,000 mental health focused economic evaluation reports were produced between 2000 to 2019 [@RN34]. The intellectual asset represented by this literature could be enhanced if many of the models described in these reports could be brought and kept up to date and made available in formats that maximised transferability to diverse decision contexts. We believe that modular and open source approaches would be well suited to accomplishing this goal and that the framework we have developed could act as an early prototype for solving some of the technical challenges of this task. Ideally such a program of research would be resourced to be sustained over the medium to long term and to engage a diverse network of investigators, contributors and advisers from high, middle and low income countries.

@@

Ideally health economists would explore these complex topics in partnership with modellers from other disciplines (in particular epidemiology and health services research) and a wide range of stakeholders such as other researchers, policymakers, service planners and community members. Modular and open source approaches would facilitate such investigations by breaking down ambitious and long term goals into manageable time-bound discrete projects, each progressed by different teams. To facilitate such an approach, a common framework of standards and tools would be needed. To be suitable for such a task our framework would need additional development, with its overall architecture reviewed for scalability and suitability and to provide better integration with and use of other open source languages (particularly python) and repositories. Whatever MOSHEM infrastructure is developed, its resilience would depend on a community of open source contributors sufficiently large and active to ensure that all core modules are maintained even after their original authors cease their involvement.

@@

Adopting MOSHEMs will expand the type of skillset typically engaged in health economic modelling projects, with a much greater role for data-scientists, software engineers and online community builders. The requirement for these roles should be incorporated into project proposals. Not all efforts by health economists to promote MOSHEMs need to depend on the decisions of research funders. Releasing selected subsets of unmaintained model artefacts in open source repositories is still better than not providing access to any code and data and can typically be accomplished within existing project budgets. Developing knowledge and skills of MOSHEMS can be advanced by making small contributions (e.g. improvements to documentation, code contributions) to open source projects. Our project website [@rfwn2022] includes details of multiple ways to contribute to ready4.

<!-- Estimating the potential uptake of these tools is difficult.  -->
<!-- Open source is more than open access (limited discussion). -->
 <!-- A coding framework for OSHEMs developed in the language R includes standardised approaches to directory structure and naming conventions [@Alarid2019]. -->
<!-- http://hdl.handle.net/1893/27205--><!-- CONTEXT: Funders should also support methodological work on context, including the development of reporting templates to facilitate accurate, comprehensive and systematic reporting of context, approaches to evidence synthesis that can capture contextual variation adequately and modelling methods that can leverage the accumulating evidence on contextual variation to predict the reach, implementation and effectiveness of promising interventions in new contexts." --

@@

<!-- Computational modelling could play an important role in developing policies to improve population mental health but this may require significant changes in the way mental health modelling projects are funded, conceptualised and implemented.  -->
 <!-- but, as with health economics more generally, OSHEMs remain rare. Greater use of open source approaches could help improve the scope, validity and usefulness of mental health economic models. -->

>>>>

<!-- We have designed this framework to enable ethical computational implementations of health economic models that are both open-source (for transparency and reusability) and modular (for reusability and updatability). The framework aims to support the standardization and interoperability that are core enablers of implementing a modular and open-source approaches.  -->

@@

Modular models provide an opportunity for multiple modelling teams to contribute to, test and reuse models. To enhance this capacity, modular models may be implemented as open-source projects that give others liberal permissions to access and use model source code and data [@pan2021modular; @barros2023empowering; @frazier2022robust]. Although there appears to be in principle support from many health economists for greater use of open-source health economic models (OSHEMs)  [@Pouwels2022], actual implementations are rare [@Emerson2019; @Michalczyk2018; @Feenstra2022]. Barriers to adoption of OSHEMs include concerns about intellectual property, confidentiality, model misuse and the resources required to support open-source implementations [@Pouwels2022; @Wu2019]. Adherence to good practice guidance is an essential requirement for healthcare modelling [@Erdemir2020], but guidelines for implementing OSHEMs remain scarce, piecemeal and need improving [@Sampson2019].

Our interest in modular and open-source approaches developed when we began seeking am appropriate framework for undertaking and validly synthesising diverse types of economic research in mental health. Mental disorders impose high health, social and economic burdens worldwide [@RN8;@GBD2019]. Much of this burden is potentially avertable [@RN25], but poorly financed and organised mental health systems are ill-equipped for this challenge [@RN22;@RN23]. A substantial economic literature already exists to assess the affordability and value for money of mental health interventions [@RN34]. This economic evaluation work is an essential prerequisite for improving allocative efficiency in mental health, but could be of greater value to systems planners if integrated with a broader program of economic research.

@@

Standardisation is an essential enabler of automation, collaboration, interoperability and transferability in modular and open-source approaches. An early requirement to implement ready4 is therefore to identify an appropriate set of standards. However, we are not aware of any consolidated source of guidance about recommended standards for MOSHEMs. Once identified, the standards specified for ready4 will need to be implemented by software, some of which will require custom development.

@@

### Standards for a reusable MOSHEM
```{r echo = F}
gpl_pc_dbl <- get_license_share("GPL")
```
To make model code and data widely re-usable by others, it is important to provide users with appropriate and explicit permissions. In the context of open-source models, there are two broad categories of licensing options. Some guidance strongly recommends the use of permissive licensing [@Wilson_2017] that provides users with great flexibility as to the purposes (including commercial) for which the content could be re-used. An alternative approach is to use copyleft licenses [@copyleft2022] that can require content users to distribute any derivative works they create under similar open-source arrangements. For code, it may be appropriate to adopt the prevailing open-source licensing practice within the programming language being used. For data, it may not be sufficient to simply choose between a permissive license like the Public Domain Dedication (CC0) [@cc02022] or a copyleft option such as the Attribution-Share Alike (CC-BY-SA) [@bysa2022]. In addition to ensuring that data is ethically appropriate for disseminate in open access repositories, responsible custodianship of some de-identified or aggregated data may involve using or adapting template terms of use [@sampleterms2022] which have a number of ethical clauses (for example, prohibiting efforts to re-identify research participants).

@@

Statistical models are a common output of health economic evaluations, but they are often not reported in a format that enables others to confidently and reliably re-use them for out of sample prediction [@Kearns2013]. open-source approaches can help address this by disseminating code artefacts that enable easy and appropriate use of a statistical model to make predictions with new data. However, great care must be exercised when publicly releasing model artefacts derived from data on human subjects as they may by default embed a copy of the source dataset. Sensitive dataset copies must therefore be replaced (for example, with synthetic data) and the amended artefact's predictive performance then retested before any public release. Another way to make MOSHEMs easier to use is to develop simple user-interfaces for non-technical users.

@@

However, having features that facilitate accountability, reuse and updating is not the same as being accountable, reused and updated. If diverse groups of stakeholders do not review model components, suggest improvements and develop alternatives, then little progress is made towards enhancing model legitimacy. Similarly, making code and data publicly available does not guarantee that others will know of the existence of these tools, trust their validity and find them easy to use. Without reuse, errors in model artefacts are more likely to remain undetected. Even when errors are detected, they still need to be fixed, but maintaining code and data requires ongoing resourcing through a combination of centralised infrastructure and an active open-source community.

@@

## Implications for economic modelling in mental health
open-source approaches have been recommended to help develop the mental health modelling field [@RN73] but only one mental health related model (in Alcohol Use Disorder [@Basu2018]) is currently indexed in the open-source Models Clearinghouse [@OSMC_20xx;@Emerson2019]. We are aware of just one other open-source mental health model - a reference model in Major Depressive Disorder - that is currently in development [@IVIMDD2022]. Of the known barriers to adoption of open-source models by health economists (including issues like intellectual property and confidentiality [@Pouwels2022; @Wu2019]), our experience suggests that the biggest challenges may be the enormous effort required to first prepare model code and data for public release in formats that facilitate appropriate reuse by third parties and to then maintain and continually improve potentially large numbers of digital artefacts.

Automated tools such as those we developed in our modelling toolkit can help reduce the burden associated with some of these tasks. However, we think the current low rates of adoption by health economists of open-source approaches [@Emerson2019; @Michalczyk2018; @Feenstra2022] (which in turn facilitate the collaboration that make modular models more attractive) will only change slowly unless there is significant and strategic investments made by research funders. Currently, incentive structures for health economists do not promote the dedication of large quantities of time to enable peers to reuse their work.<!-- MH Research priorities (WP3 - Research, WP6 - Economic) https://doi.org/10.1016/S2215-0366(15)00332-6 -->

Reducing waste in research is a responsibility of research funders [@chalmers2014increase] and the poor reproducibility [@Jalali2021; @McManus2019; @Bermejo2017], limited reusability [@Emerson2019; @Michalczyk2018; @Feenstra2022] and uncertain validity [@Ghabri2019; @Radeva2020] of health economic models is wasteful. Approximately 4,000 mental health focused economic evaluation reports were produced between 2000 to 2019 [@RN34]. The intellectual asset represented by this literature could be enhanced if many of the models described in these reports could be brought and kept up to date and made available in formats that maximised transferability to diverse decision contexts. We believe that modular and open-source approaches would be well suited to accomplishing this goal and that the framework we have developed could act as an early prototype for solving some of the technical challenges of this task. Ideally such a program of research would be resourced to be sustained over the medium to long term and to engage a diverse network of investigators, contributors and advisers from high, middle and low income countries.

@@

Ideally health economists would explore these complex topics in partnership with modellers from other disciplines (in particular epidemiology and health services research) and a wide range of stakeholders such as other researchers, policymakers, service planners and community members. Modular and open-source approaches would facilitate such investigations by breaking down ambitious and long term goals into manageable time-bound discrete projects, each progressed by different teams. To facilitate such an approach, a common framework of standards and tools would be needed. To be suitable for such a task our framework would need additional development, with its overall architecture reviewed for scalability and suitability and to provide better integration with and use of other open-source languages (particularly python) and repositories. Whatever MOSHEM infrastructure is developed, its resilience would depend on a community of open-source contributors sufficiently large and active to ensure that all core modules are maintained even after their original authors cease their involvement.

@@

Adopting MOSHEMs will expand the type of skillset typically engaged in health economic modelling projects, with a much greater role for data-scientists, software engineers and online community builders. The requirement for these roles should be incorporated into project proposals. Not all efforts by health economists to promote MOSHEMs need to depend on the decisions of research funders. Releasing selected subsets of unmaintained model artefacts in open-source repositories is still better than not providing access to any code and data and can typically be accomplished within existing project budgets. Developing knowledge and skills of MOSHEMS can be advanced by making small contributions (e.g. improvements to documentation, code contributions) to open-source projects. Our project website [@rfwn2022] includes details of multiple ways to contribute to ready4.

<!-- Estimating the potential uptake of these tools is difficult.  -->
<!-- open-source is more than open access (limited discussion). -->
 <!-- A coding framework for OSHEMs developed in the language R includes standardised approaches to directory structure and naming conventions [@Alarid2019]. -->
<!-- http://hdl.handle.net/1893/27205--><!-- CONTEXT: Funders should also support methodological work on context, including the development of reporting templates to facilitate accurate, comprehensive and systematic reporting of context, approaches to evidence synthesis that can capture contextual variation adequately and modelling methods that can leverage the accumulating evidence on contextual variation to predict the reach, implementation and effectiveness of promising interventions in new contexts." --

@@

<!-- Computational modelling could play an important role in developing policies to improve population mental health but this may require significant changes in the way mental health modelling projects are funded, conceptualised and implemented.  -->
 <!-- but, as with health economics more generally, OSHEMs remain rare. Greater use of open-source approaches could help improve the scope, validity and usefulness of mental health economic models. -->

### 2024-03-19 Markdown/Manuscript/Child_RMDs/MS_Aims.Rmd
v4

---
title: "MS_Aims"
author: "Author"
date: "`r Sys.Date()`"
output: html_document
---

In this paper, we <!--  MOVED_EDITED_START -->consider literature on modelling practice and draw on our prior experience with developing and using CHEMs to propose<!--  MOVED_EDITED_START -->:  (i) a set of ethical responsibilities for CHEM project teams<!-- and criteria for assessing ethical CHEM implementations -->;<!--  MOVED_EDITED_START --> (ii) attributes of CHEMs that can suggest fulfillment of these responsibilities; and (iii) criteria against which these attributes can be assessed.  <!--  MOVED_EDITED_END -->  
<!-- (i) some core ethical responsibilities of CHEM project teams;

>>>>

---
title: "MS_Aims"
author: "Author"
date: "`r Sys.Date()`"
output: html_document
---
<!-- Important rewordings --> 
In this paper, we <!--  MOVED_EDITED_START -->draw on our prior experience with developing and using CHEMs and consider literature on modelling practice to propose<!--  MOVED_EDITED_END -->:  (i) a set of research ethics <!--  EDITED_Start responsibilities-->goals<!--  _EDITED_END --> for CHEM <!--  EDIT -->developers and funders<!--  /EDIT --<!-- and criteria for assessing ethical CHEM implementations -->;<!--  MOVED_EDITED_START --> (ii) attributes of CHEMs that suggest achievement of these goals; and (iii) criteria against which these attributes can be assessed.  <!--  MOVED_EDITED_END -->  
<!-- (i) some core ethical responsibilities of CHEM project teams;

### 2024-03-19 Markdown/Manuscript/Child_RMDs/MS_Cndl_Tabs_Not_In_Body.Rmd
v4

<!-- # ```{r, child = '../Child_RMDs/MS_Tables_3.Rmd', echo=FALSE} -->
<!-- # ``` -->
<!-- # \newpage -->
<!-- # ```{r, child = '../Child_RMDs/MS_Tables_4.Rmd', echo=FALSE} -->
<!-- # ``` -->

>>>>

```{r, child = '../Child_RMDs/MS_Tables_3.Rmd', echo=FALSE}
```
\newpage
<!-- # ```{r, child = '../Child_RMDs/MS_Tables_4.Rmd', echo=FALSE} -->
<!-- # ``` -->

### 2024-03-19 Markdown/Manuscript/Child_RMDs/MS_Conclusion.Rmd
v4

---
title: "MS_Conclusion"
author: "Author"
date: "`r Sys.Date()`"
output: html_document
---
`r ifelse(is.na(params$X@conclusion_1L_chr),"",params$X@conclusion_1L_chr)` 
<!-- We have identified criteria that can be used to systematically assess the extent to which the computational implementation of health economic models adheres to transparency, reusability and updatability criteria. <!-- We have developed a prototype software framework that can support the implementation of CHEMs that largely meet these criteria. Our framework can be used as a prototype for developing future software frameworks to support ethical implementation of CHEMs.  --> 
<!-- Barriers to ethical CHEM practice arising from resource constraints should be addressed by research funders. -->

>>>>

---
title: "MS_Conclusion"
author: "Author"
date: "`r Sys.Date()`"
output: html_document
---
`r ifelse(is.na(params$X@conclusion_1L_chr),"",params$X@conclusion_1L_chr)` 
<!-- We have identified criteria that can be used to systematically assess the extent to which the computational implementation of health economic models adheres to transparency, reusability and updatability criteria. <!-- We have developed a prototype software framework that can support the implementation of CHEMs that largely meet these criteria. Our framework can be used as a prototype for developing future software frameworks to support ethical implementation of CHEMs.  --> 
<!-- Barriers to ethical CHEM practice arising from resource constraints should be addressed by research funders. -->
 We propose that CHEM developers and funders adopt the ethical goals of socially acceptable user requirements and design specifications, fit for purpose implementations and socially beneficial post-release use. We suggest that CHEM attributes of transparency, reusability and updatability are indicative of projects that meet these goals and have identified six criteria for assessing these attributes. We hope our proposals contribute to a timely and much needed debate within the health economics community about the ethical dimensions of implementing health economic models as software.

### 2024-03-19 Markdown/Manuscript/Child_RMDs/MS_Discussion.Rmd
v4

Ethical practice is a core expectation of health researchers and computational methods underpin most quantitative research. However, an understanding of the ethical dimensions about how health economic models are implemented as software is underdeveloped. We have attempted to address this gap by proposing that CHEM developers and funders adopt the ethical goals of socially acceptable user requirements and design specifications, fit for purpose implementations and socially beneficial post-release use. We suggest that CHEM attributes of transparency, reusability and updatability are indicative of projects that meet these goals and have identified six criteria for assessing these attributes.

Prior literature has noted that health economic models routinely have limitations in transparency [@Jalali2021; @McManus2019; @Bermejo2017; @Ghabri2019], reuse [@Feenstra2022; @Emerson2019], and being kept up to date [@Sampson2017; @10.3389/fpubh.2022.899874]. These limitations not only pose ethical risks such as those we have described, but are practical barriers to model generalisability and transferability. Addressing these limitations will require changes to the work practices, funding and tools that underpin health economic model development.

Adherence to good practice guidance is an essential requirement for healthcare modelling [@Erdemir2020] and there is a need for better and comprehensive guidelines about how health economists should implement health economic models as software. Research funders need to actively support initiatives to help remove barriers to better practice in this area. A core responsibility of research funders is reducing waste in research [@chalmers2014increase]. Investments in modelling projects might generate greater social returns if they supported the project activities necessary to develop CHEMs that are easily understood, widely reused and well maintained.

Previously recommended strategies for more beneficial health economic research investments include support for harmonised ethical standards for model development [@10.3389/fpubh.2017.00068], methodological innovation to improve model transferability [@craig2018taking], networks of modellers working on common health conditions [@Sampson2019], and  centralized infrastructure such as open source model repositories [@Pouwels2022] and a standard platform for model implementations [@Ghabri2019]. Each of these strategies could benefit from enabling technical infrastructure, for example software frameworks that support the development of CHEMs that meet ethical criteria. We are currently engaged in developing a prototype of one such software framework [@ready4fw2024].

@@

<!-- However, our prototype has a number of features that future work on CHEM software frameworks may find useful to incorporate. Firstly, developing a software framework to work within an existing and widely used open source programming language such as R, can keep framework scope relatively narrow. This makes it more tractable to develop, maintain and learn, and facilitates integration with other modelling tools written in that language (e.g., the dependency libraries we list in Table \@ref(tab:cpkgs)). Secondly, implementation that combines both object-oriented and functional programming paradigms can avail of the modular and syntactical simplicity benefits of the former, while limiting needless bundling of code artefacts. Thirdly, a sensible trade-off needs to be found between transparent code implementation (which requires clear and sufficiently detailed documentation) and Agile Software Development (for which a foundational principle is prioritising the development of working code over writing documentation [@beck2001manifesto]). Our software framework makes this trade off by enforcing the use of consistent code style conventions and file organisation which in turn enables automated generation of simple documentation at every CHEM module update. All model data-structures and algorithms are therefore always documented (at least minimally, with machine authored content), meaning model developers have a requirement to write customised documentation less frequently. Many of the features of our prototype software framework have the potential to be significantly enhanced through integration with evolving code and documentation authoring capabilities of large language models. -->

<!-- Our software framework provides the outline of a technical solution to some of the challenges of open source CHEM development. However, better tools alone are insufficient and policy and institutional changes are also required.  -->

>>>>

Prior literature has noted that health economic models routinely have limitations in transparency [@Jalali2021; @McManus2019; @Bermejo2017; @Ghabri2019], reuse [@Feenstra2022; @Emerson2019], and being kept up to date [@Sampson2017; @10.3389/fpubh.2022.899874]. These limitations not only pose ethical risks such as those we have described, but are practical barriers to achieving potential exciting enhancements to health economic modelling practice such as improved model transferability [@munthe2019systematic; @barbieri2010international], greater use of reference models [@Afzali2013], living health technology assessments [@thokala2023living] and leveraging the analytic capabilities of machine learning [@padula2022machine].

It is important that the health economics community engages with the ethical dimensions of how health economic models are implemented as software. Recent developments in Large Language Models (LLMs) mean software developers are confronted with novel ethical issues including: increased risk of security flaws; a potential need to add restrictions to open-source licensing; environmentally harmful levels of computing power; and concerns about model explainability and unintended consequences [@ozkaya2023application]. Early work exploring the use of LLMs in CHEM development has already been undertaken [@poirrier2023msr26]. Many of the active ethics debates in the computer science community are therefore likely to become increasing relevant to health economists and public health researchers. Documents such as the World Health Organisation's recent guidance on generative artificial intelligence [@world2024ethics] will be important reference points for how the health economics discipline navigates emerging ethical themes in CHEM development.

Meeting all six TRU criteria we propose is more likely when a CHEM has been implemented as an open-source project<!-- and model transferability -->. It should be noted that the ethical dimensions of open-source software development are broader than the issues we have discussed here, including issues relating to social contracts, contributor autonomy and public goods [@grodzinsky2003ethical]. There are also important barriers to growing the (currently paltry) number of open-source CHEMs. These barriers include health economist concerns about intellectual property, confidential data, the risk of model misuse and the resources required for model maintenance [@Pouwels2022; @Wu2019]. Commercial considerations, for example when health economic models are owned by pharmaceutical companies and consultancies, may also limit the public availability of models and their constituent code and data [@Feenstra2022]. Other barriers relate to <!-- model generalisability and transferability and  -->the software platforms in which CHEMs are authored [@Pouwels2022].

Recent efforts to develop the open-source CHEMs field include an initiative to develop a suite of open source CHEMs addressing different health conditions in the United States [@Jansen2019] and a Canadian platform facilitating access to and validation of multiple CHEMs [@harvard2022developing]. It has also been recommended that training curricula for current and future health economists need to develop some foundational software development skills [@kindilien2021role]. Education efforts to develop such skills could be reinforced with best-practice scientific computing guidance for health economists. Adherence to good practice guidance is an essential requirement for healthcare modelling [@Erdemir2020], but guidelines for implementing open source CHEMs remain scarce, piecemeal and need improving [@Sampson2019]. Health economists also need to be able to access reliable and up to date legal advice about the implications of alternative approaches to software licensing.

<!-- Move to discussion--><!-- In health economics the related and enabling concept of reference models has been recommended [@Afzali2013].  -->
<!-- Transferring a health economic model developed for one jurisdiction for application in another typically involves retaining some features and updating others [@barbieri2010international]. However, models do not always make implementation choices (for example, use of concepts with standardised meanings across jurisdictions)<!-- [@garcia2021cost] -->  <!-- facilitate such selective re-use. -->
<!-- Open-source CHEMs remain rare [@Feenstra2022; @Emerson2019]. The software used to develop CHEMs was identified as a barrier to open-source model implementations by participants in a 2020 survey of health economists [@Pouwels2022]. Other reported barriers relate to difficulties in generalising models, updating models and transferring model data, concerns about the level of public access and lack of interest from decision-makers, and legal, confidentiality and security considerations [@Pouwels2022].  -->

Many of these barriers require significant investments by governments and other research funders in enabling infrastructure and human capital. <!-- A core responsibility of research funders is reducing waste in research [@chalmers2014increase]. Investments in modelling projects might generate greater social returns if they supported the project activities necessary to develop CHEMs that are easily understood, widely reused and well maintained.  --> Investment strategies that have previously been recommended for development of the health economic research field include support for harmonised ethical standards for model development [@10.3389/fpubh.2017.00068], methodological innovation to improve model transferability [@craig2018taking], networks of modellers working on common health conditions [@Sampson2019], and  centralized infrastructure such as open-source model repositories [@Pouwels2022] and a standard platform for model implementations [@Ghabri2019].

<!-- Each of these strategies could benefit from enabling technical infrastructure, for example software frameworks that support the development of CHEMs that meet ethical criteria. We are currently engaged in developing a prototype of one such software framework [@ready4fw2024]. -->

<!-- Adherence to good practice guidance is an essential requirement for healthcare modelling [@Erdemir2020] and there is a need for better and comprehensive guidelines about how health economists should implement health economic models as software.  -->

@@

<!-- However, our prototype has a number of features that future work on CHEM software frameworks may find useful to incorporate. Firstly, developing a software framework to work within an existing and widely used open-source programming language such as R, can keep framework scope relatively narrow. This makes it more tractable to develop, maintain and learn, and facilitates integration with other modelling tools written in that language (e.g., the dependency libraries we list in Table \@ref(tab:cpkgs)). Secondly, implementation that combines both object-oriented and functional programming paradigms can avail of the modular and syntactical simplicity benefits of the former, while limiting needless bundling of code artefacts. Thirdly, a sensible trade-off needs to be found between transparent code implementation (which requires clear and sufficiently detailed documentation) and Agile Software Development (for which a foundational principle is prioritising the development of working code over writing documentation [@beck2001manifesto]). Our software framework makes this trade off by enforcing the use of consistent code style conventions and file organisation which in turn enables automated generation of simple documentation at every CHEM module update. All model data-structures and algorithms are therefore always documented (at least minimally, with machine authored content), meaning model developers have a requirement to write customised documentation less frequently. Many of the features of our prototype software framework have the potential to be significantly enhanced through integration with evolving code and documentation authoring capabilities of large language models. -->

<!-- Our software framework provides the outline of a technical solution to some of the challenges of open-source CHEM development. However, better tools alone are insufficient and policy and institutional changes are also required.  -->
